---
title: "Creating raster mask from shapefile"
author: "Denisse Fierro Arcos"
date: "2023-05-08"
output: 
  github_document:
    toc: true
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
In this notebook, we will a shapefile from [VLIZ](https://doi.org/10.14284/386) which contains the boundaries of all the Exclusive Economic Zones (EEZs) around the world to create a raster mask for 22 EEZs of our interest. All these EEZ are located within the South Pacific region and are listed in this [website](https://stats.pacificdata.org/vis?lc=en&df[ds]=SPC2&df[id]=DF_KEYFACTS&df[ag]=SPC&df[vs]=1.0&dq=..EEZ.&pd=2020%2C2021&ly[rw]=GEO_PICT). We will access the South Pacific EEZ list remotely.  
  
The resulting raster mask for the South Pacific EEZs will be used to extract ISIMIP3B input and output data that will be used to perform vulnerability assessments of this region. This notebook will use a combination of `R` and `Python` with the `reticulate` library.

## Loading R libraries

```{r libraries, results = "hide", warnings = F, message = F}
library(sf)
library(raster)
library(tidyverse)
library(reticulate)
```

## Loading EEZs of the world shapefile
This shapefile was downloaded from [VLIZ](https://doi.org/10.14284/386) and it is not included in this repository due to its size. Ensure you have downloaded this file and saved it inside the `Data` folder.  
  
**Note:** The shapefile version 11 downloaded on November 9, 2022 is used in this notebook. If a later version is available, ensure the file paths to the shapefile is updated in the code chunk below.

```{r}
eez_world <- read_sf("../Data/World_EEZ_v11_20191118/eez_v11.shp") %>% 
  #We only need a subset of all columns available in the original dataset
  select(MRGID, GEONAME, SOVEREIGN1, SOVEREIGN2, SOVEREIGN3, TERRITORY1, TERRITORY2, TERRITORY3, AREA_KM2) #%>%
  # #Turning character columns into factors
  # mutate_if(is.character, as.factor)
```

## Loading list of EEZs within the South Pacific region
As mentioned in the introduction, we will only extract data for 22 EEZs located within the South Pacific region. A list of all relevant EEZs in this region can be found in the [Pacific Data Hub](https://pacificdata.org/). We will access this list through an API.
```{r}
#Accessing data
SP_EEZ <- jsonlite::fromJSON("https://stats-sdmx-disseminate.pacificdata.org/rest/data/SPC,DF_KEYFACTS,1.0/..EEZ.?startPeriod=2020&endPeriod=2021&dimensionAtObservation=AllDimensions", simplifyDataFrame = T)

#Extracting EEZ list
SP_EEZ <- SP_EEZ$structure$dimensions$observation %>% 
  filter(str_detect(name, "Pacific")) %>% 
  select(values) %>% 
  pull(values) %>% 
  #Converting to data frame
  as.data.frame() %>% 
  #Removing names given in parenthesis before matching to EEZ shapefile
  mutate(name = str_remove(name, " \\(.*\\)"))

head(SP_EEZ)
```

## Filtering South Pacific EEZs
We will do a `left_join` of the EEZ list above to the `TERRITORY1` column in our EEZ shapefile. This will extract most of the EEZs of our interest. This will be followed by a partial match using the `GEONAME` column for any EEZs missing. We will merge these results into a single variable.

```{r}
#Initial partial match
eez_SP <- SP_EEZ %>% 
  left_join(eez_world, by = join_by("name" == "TERRITORY1"), keep = T)

#Checking missing EEZs from initial match
missing_EEZ <- eez_SP %>% 
  filter(is.na(MRGID)) %>% 
  select(id, name)

missing_EEZ
```
We will now perform a second partial search using the `GEONAME` column of the EEZ shapefile, and merge it to our first match results.

```{r}
#Performing a partial search and binding results together
eez_SP <-  eez_world %>% 
  filter(str_detect(GEONAME, paste(missing_EEZ$name, collapse = "|"))) %>% 
  #Adding South Pacific EEZ names
  bind_cols(missing_EEZ, .) %>% 
  #Merging to initial results
  bind_rows(eez_SP) %>% 
  #Dropping rows with no results
  drop_na(MRGID) %>% 
  #Turning character columns into factors
  mutate_if(is.character, as.factor) %>% 
  #Converting back to shapefile
  st_as_sf()
  
#Checking all South Pacific EEZs are included in final result
eez_SP$name %in% SP_EEZ$name
```
We will plot the shapefile to inspect the results.
```{r}
#Loading land shapefile to include in plot
land <- rnaturalearth::ne_countries(type = "countries", returnclass = "sf")

#Plotting FAO shapefile
eez_SP %>% 
  ggplot()+
  geom_sf(aes(fill = name))+
  geom_sf(data = land, inherit.aes = F, color = "gray")+
  theme_bw()+
  theme(legend.position = "none")+
  lims(y = c(-40, 30))
```
All the EEZs are located within our area of interest, the South Pacific, we can now save the results.

## Saving results
We will extract the EEZ names and unique numeric IDs to create the raster masks.

```{r}
#Identify EEZ and countries associated with them from extracted data files
eez_SP_codes <- eez_SP %>% 
  select(name, MRGID) %>% 
  st_drop_geometry()

#Saving results
eez_SP_codes %>% 
  write_csv("../Outputs/SouthPacific_EEZ_keys.csv")
```

We will save the shapefile with the South Pacific EEZs.

```{r}
eez_SP %>% 
  #We will also allow shapefile to be rewritten
  st_write("../Outputs/SouthPacific_EEZ.shp", delete_layer = T)
```
## Extracting names and codes for EEZ
We will use this information to save correct names for regions in the raster

```{r eez_keys}
#Create a data frame of unique EEZs
eez_SP_codes <- eez_SP_codes %>% 
  group_by(name) %>% 
  group_keys()

#We can check some of the results
head(eez_SP_codes, 2)
```

## Creating a multilayer raster mask based on merged shapefile
We will now create a multilayer mask that matches the grid used in the physical model forcings. Most models use the same 1 degree grid, with the exception of the DBPM ecosystem model, which uses a different 1 degree grid, and the DBEM model, which uses a 0.5 degree grid.  
  
Note that masks *must* match the grid of the model from which data is being extracted. This means that you will need to create a new mask for each grid that is different. In the chunk below, you will find the three different grids identified in the ecosystem models.

### Loading input rasters

```{r load_rasters_input}
#Loading sample raster to be used as target for rasterising FAO regions
#Most ecosystem models use this one degree grid
ras <- raster("../Spatial_Data/InputRasters/gfdl-mom6-cobalt2_obsclim_deptho_1deg_global_fixed.nc")

#Sample from DBPM model
# ras <- raster("../Spatial_Data/InputRasters/dbpm_ipsl-cm6a-lr_nobasd_historical_nat_default_1deg_tcb_global_monthly_1850_2014.nc")[[1]]

#Sample from DBEM model
# ras <- raster("../Spatial_Data/InputRasters/dbem_ipsl-cm6a-lr_nobasd_historical_nat_default_05deg_tcb_global_annual_1951_2014.nc")[[1]]

#Sample for 0.25 degree grid
# ras <- raster("../Spatial_Data/InputRasters/gfdl-mom6-cobalt2_obsclim_deptho_025deg_global_fixed.nc")

#We will define a few extra variables to automate creation of file names for each mask
#Model resolution
res <- "1deg"

#Model associated with grid. Leave blank if multiple models use same grid
mod_name <- ""

#Plotting raster
plot(ras)
```
### Calculate grid area
The `raster` package allows us to calculate the area of grid cells in $km^2$ in just one line of code. We can save this to calculate weighted means by area. We will use the raster above as a base, and we will save the result on our disk.

```{r eval = F}
#Calculating grid area
area_grid <- area(ras)

#Creating file name 
filename <- paste0("area_", res, mod_name, ".nc")

#Saving results
writeRaster(area_grid, file.path("../Spatial_Data/Masks", filename), format = "CDF", overwrite = T, 
            #We will save the variable names so they match the Fish-MIP models
            varname = "area", xname = "lon", yname = "lat", varunit = "km2")
```

### Defining function to create rasters from shapefiles
We will define our own function that will use the shapefiles above to create rasters.

```{r function_shp_raster}
#Defining function which needs a shapefile and a raster as input
shp_to_raster <- function(shp, nc_raster){
  #The final raster will have ones where within the shapefile boundaries
  rasterize(shp, nc_raster, field = 1)
}

```

### Applying raster creation function
We will apply the function we defined above to all the polygons contained within our shapefile. We can do this easily using the `map` function from the `purrr` package.

```{r mapping_function}
#Split shapefile into regions prior to transforming into raster
eez_list <- eez_SP %>% 
  group_by(name) %>% 
  group_split()

#Applying function to raster list
grid_raster <- map(eez_list, shp_to_raster, ras) %>% 
  #Stacking rasters to create a single multilayer raster
  stack()

#Checking results of stacked raster (selected Australian EEZs)...
plot(grid_raster[[c(1:4)]])
```

Finally, we save the outputs in our local machine.
```{r, eval = F}
#Define file name
filename <- paste0("EEZ-world_", res, "mask", mod_name, ".nc")

#Saving raster to disk
writeRaster(grid_raster, file.path("../Spatial_Data/Masks", filename), format = "CDF", overwrite = T,
            varname = "EEZ_regions", zname = "Country_EEZ")
```

# `Python`-based code
We will now start `Python` and save the correct names for the FAO regions in the `netcdf` file we created in `R`.

```{r}
#Activating conda
use_condaenv("CMIP6_data")
```

## Loading libraries

```{python}
import xarray as xr
import numpy as np
import matplotlib.pyplot as plt
import pandas as pd
import os
```

## Loading raster using `xarray`
```{python}
#Getting filename
fn = f'EEZ-world_{r.res}mask{r.mod_name}.nc'

#Loading multilayer raster as dataset
mask = xr.open_dataset(os.path.join('../Spatial_Data/Masks', fn))
#Checking saved file in R
mask
```
Here we can see that the FAO region names are not saved correctly. They are numbered based on its location on the shapefile. We can update this using the data frame with the unique FAO regions we previously created.

```{python}
#We can simply load the data frame from the R environment.
eez_keys = r.eez_names_codes

#Checking list
eez_keys
```
We can now update the names on the `netcdf` file using the ID codes for each EEZ.

```{python}
#We will use the values in the data frame we loaded above
mask['Country_EEZ'] = [int(i) for i in eez_keys.MRGID]
```

We will load the full keys into memory and we will select all EEZ areas for Australia to ensure masks are correct.

```{python}
#Create filename 
#Loading keys
eez_full_keys = pd.read_csv('../Spatial_Data/World_EEZ_v11_20191118/eez_key.csv')

#Selecting Australian subset
aus_eez = mask.sel(Country_EEZ = eez_full_keys['MRGID'][eez_full_keys['SOVEREIGN1'] == 'Australia'].tolist()).EEZ_regions
```


## Plotting results
We will plot all regions below to ensure we got them all correctly.
```{python}
#We will loop through the first five to check results
for reg in aus_eez:
  #Plotting results
  fig = plt.figure()
  ax = fig.add_subplot(111)
  reg.plot(ax = ax, levels = [1, 2])
  plt.title(reg.Country_EEZ.values.tolist())
  plt.show()

```

When comparing to the shapefile plots in the `R` section of this notebook, we can see that the regions are named correctly. This means that we can save our results now.

```{python, eval = F}
#Creating filename
fn = f'EEZ-world-corrected_{r.res}mask{r.mod_name}.nc'

mask.to_netcdf(os.path.join('../Spatial_Data/Masks', fn))
```

## Creating a 2D mask in `csv` format
Depending on the format of your data, it may be better to have a 2D mask in `csv` format. We can create one using the raster we created in the previous step. We will switch to `R` to complete this step.

```{r}
#First we load our raster mask
filename <- paste0("EEZ-world-corrected_", res, "mask", mod_name, ".nc")
# We will load it as a stack
ras <- stack(file.path("../Spatial_Data/Masks", filename), 
             #This is the same variable name we used to create our original raster
             varname = "EEZ_regions")

#We will now convert our stack into points
ras_df <- rasterToPoints(ras) %>% 
  #Transforming form matrix to data frame
  as.data.frame() %>% 
  #Changing dimensions of data frame, so that our column names appear in the `ID_merged column`
  pivot_longer(cols = -c(x, y), names_to = "ID_merged") %>% 
  #We remove any rows containing only NA values
  drop_na(value) %>% 
  #We rename latitude and longitude column
  rename("Lon" = "x", "Lat" = "y") %>% 
  #We remove the `X` from the areas IDs
  mutate(ID_merged = as.integer(str_remove(ID_merged, 'X'))) %>% 
  #Remove the `value` column as it is now redundant
  select(-value)

#We can update our filename variable 
filename <- paste0("EEZ-world_", res, "mask", mod_name, ".csv")
#Saving the end file in your local machine
write_csv(ras_df, file = file.path("../Spatial_Data/Masks", filename))
```
