---
title: "Biomass projections for Pacific Islands Countries and Territories (PICTs)"
author: "Beth Fulton and Denisse Fierro Arcos"
date: "2023-11-29"
output:
  github_document:
    toc: true
    html_preview: false
---

# Marine biomass projections for Pacific Islands Countries and Territories (PICTs)
The workflow described in this notebook was developed by [Dr Beth Fulton](https://orcid.org/0000-0002-5904-7917) (CSIRO) and implemented in `R` by [Denisse Fierro Arcos](https://github.com/lidefi87/).  
  
The [Fisheries and Marine Ecosystem Model Intercomparison Project (FishMIP)](https://fish-mip.github.io/) includes global marine ecosystem models that vary in their complexity and the equations they used to capture relationships between the abiotic and biotic components of marine ecosystems. FishMIP provide estimates of biomass of a wide range of marine organisms, including fish and invertebrates, from 1965 to 2100. FishMIP models are forced by environmental data from two earth system models (ESM): GFDL-ESM4.1 and IPSL-CM6A-LR. The ESM inputs used come from the historical run (1965-2014) and two emissions scenarios: SSP1-2.6 (low emissions) and SSP5-8.5 (high emissions).  
  
These FishMIP models are global in scope, and so they may not truly capture the relationships between coral cover and fish biomass. In this notebook, we will improve FishMIP biomass projections by taking into account the effect of coral via a bias correction, which is explained in detailed in the sections below.  
  
Finally, we will calculate an ensemble mean using the corrected fish biomass values from all available FishMIP models.   
  
## Loading relevant libraries
  
```{r, message = F, warning = F}
library(tidyverse)
library(openxlsx)
library(janitor)
```
    
##  Models used to generate biomass projections for PICTs
A brief description of the six FishMIP models used in our ensemble is included below, and you can click on the links to find out more information about each model.  
1. [APECOSM](https://apecosm.org/): Apex Predators ECOSystem Model, which represents the spatialized dynamics of open ocean pelagic ecosystems in the global ocean.  
2. [BOATS](https://earthsystemdynamics.org/models/boats/): BiOeconomic mArine Trophic Size-spectrum model simulates the global fishery as a coupled ecological-economic system.  
3. [DBPM](https://bitly.ws/36wjv): The Dynamic Benthic Pelagic Model is a dynamic size spectrum model for modelling the coupling "pelagic" size-based predators and "benthic" detritivores that share a unstructured resource pool (detritus).  
4. [EcoTroph](https://doi.org/10.1016/j.ecolmodel.2009.07.031): EcoTroph models the functioning of marine ecosystems as flows of biomass from low to high trophic levels, so as to quantify easily the impacts of fishing at an ecosystem scale.  
5. [Macroecological](https://doi.org/10.1371/journal.pone.0133794): Macroecological is a static equilibrium model, which uses ecological and metabolic scaling theory to predict mean size composition and abundance of animals (including fish).  
6. [ZooMSS](https://doi.org/10.1016/j.ecolmodel.2020.109265): The Zooplankton Model of Size Spectra is a functional size-spectrum model of the marine ecosystem to resolve phytoplankton, nine (9) zooplankton functional groups (heterotrophic flagellates and ciliates, omnivorous and carnivorous copepods, larvaceans, euphausiids, salps, chaetognaths and jellyfish) and three size-based fish groups.  
  
## Calculating bias correction for FishMIP biomass data
Bias corrections were derived from REEFMOD data, which describes the relationship between coral cover and fish biomass in the Great Barrier Reef (GBR).  

### 1. Calculating fish biomass from REEFMOD coral cover data
A quick synthesis of existing relationships can be found in the document `Relationships for fish biomass model.docx`. The most useful relationships between coral cover and fish biomass come from the [Graham and Nash (2013)](https://doi.org/10.1007/s00338-012-0984-y) data set (explored in `graham_nash_2012_dataset.xlsx`).  
  
Using the raw data from REEFMOD, we obtained the following relationships:

**Equation 1**: Structural complexity as a function of coral cover. ($r^2$ = 0.4314)  
$struct_{complexity} = 3e^{-7} \times {coral_{cover}}^4 - 5e^{-5} \times {coral_{cover}}^3 + 0.0022 \times {coral_{cover}}^2 + 1.3892$  
  
```{r}
#Base spreadsheet
working_data <- "../Analysis/Excel Sheets/bias_correction_calculations_mean_coral_cover_all_climate_scenarios.xlsx"

#Loading coral cover data
coral <- read.xlsx(working_data, sheet = 1)

#Calculating structural complexity using equation 1 above
coral <- coral |> 
  #rearrange data to facilitate calculations
  pivot_longer(-Year, names_to = "scenario", values_to = "coral_cover") |> 
  #make names small case
  clean_names() |> 
  #calculate structural complexity
  mutate(struct_complex = (3e-7*(coral_cover^4))-((5e-5)*(coral_cover^3))+(0.0022*(coral_cover^2))+1.3892)

#Checking results
head(coral)
```
  
**Equation 2**: Fish biomass ($kg \times ha^{-1}$) as a function of structural complexity ($r^2$ = 0.6102). Note that a logistic is a little poorer fit, so using a quadratic on the argument that once a reef habitat is too complex, it loses places for fish to sit.  
$fish_{biomass} = -2294.6 \times {struct_{complexity}}^2 + 8961.1 \times struct_{complexity} - 6843.6$  
  
In this step, we will also convert REEFMOD fish biomass from $kg \times ha^{-1}$ to $g \times m^{-2}$, so it matches the outputs of FishMIP models.  
  
```{r}
#Calculating fish biomass
coral <- coral |> 
  mutate(fish_biomass_kg_ha = (-2294.6*(struct_complex^2))+(8961.1*struct_complex)-6843.6) |> 
  mutate(fish_biomass = fish_biomass_kg_ha*1000/10000)

#Checking results
head(coral)
```
  
The REEF_MOD project (of Yves-Marie Bozec and Peter Mumby at University of Queensland) has postulated a slightly different set of relationships from the same data (basically a linear relationship).  
  
Fitting a line to the lower, median and upper bounds of this relationship gives the following equations:  
  
**Equation 3**: REEFMOD lower bound equation: $fish_{biomass} = 12.716 \times coral_{cover} + 146.75$  
**Equation 4**: REEFMOD median equation: $fish_{biomass} = 13.56 \times coral_{cover} + 732.15$  
**Equation 5**: REEFMOD upper bound equation: $fish_{biomass} = 14.285 \times coral_{cover} + 1325.7$  
  
```{r}
#Calculating REEFMOD biomass values (in g/m2)
coral <- coral |> 
  #lower biomass bound
  mutate(reefmod_lower_biomass = (12.716*coral_cover+146.75)*1000/10000,
         #median biomass
         reefmod_median_biomass = (13.56*coral_cover+732.15)*1000/10000,
         #upper biomass bound
         reefmod_upper_biomass = (14.285*coral_cover+1325.7)*1000/10000) |> 
  #We will drop the biomass in Kg/ha
  select(!fish_biomass_kg_ha)

head(coral)
```
  
Note that the REEFMOD team predicted mean coral coverage from 2024 to 2100 (`year` column) for the Great Barrier Reef (GBR) under five emissions scenarios: `SSP1-1.9`, `SSP1-2.6`, `SSP2-4.5`, `SSP3-7.0`, and `SSP5-8.5` (`scenario` column). Fish biomass was calculated for the entire period (2024-2100) and for all five projections.    
  
### 2. Calculating annual mean biomass estimates from FishMIP models
Here, we used the monthly biomass estimates for the historical period (1985-2014) and two emissions scenarios `SSP1-2.6` and `SSP5-8.5` (2015-2100) from [six (6) FishMIP global models](#fish-mip-models-used-to-generate-biomass-projections-for-picts) to calculate mean annual biomass estimates for each FishMIP model. We used annual mean values to match the data available from REEFMOD.    
  
**Note:** To run the code below, you will need access to the monthly biomass estimates available in the FishMIP server. If you do not have access to the server, we are providing the annual means in the code chunk.  
  
```{r, eval = F}
#Folder containing outputs from FishMIP models
base_folder <- file.path("/rd/gem/private/users/camillan/Extract_tcblog10_Data",
                         "Output/sumSize_annual/sizeConsidered10g_10kg",
                         "EEZsummaries/gridded_outputs")
#Listing all relevant files to calculate biomass projections
global_files <- list.files(base_folder, full.names = T)

#Loading PICTs EEZ mask and GBR boundaries 
mask <- read_csv("../Outputs/mask_1deg.csv")

#Defining function to extract biomass data for all PICTs from FishMIP outputs
mean_yr_bio <- function(file_name){
  da <- read_csv(file_name, col_select = x:GEONAME) |> 
    #Select data from 1985
    filter(year >= 1985) |> 
    #Rename coordinates
    rename(Lon = x, Lat = y) |> 
    #Extract GBR data - by applying mask
    right_join(mask, by = c("Lon", "Lat")) |> 
    #Calculating mean yearly biomass
    group_by(year, mem, esm, scenario, mask) |> 
    summarise(mean_annual_bio = mean(biomass, na.rm = T))
}

#Apply function to all FishMIP output files
bio_picts <- global_files |> 
  map(\(x) mean_yr_bio(x))

#Combine all list elements into a single data frame
bio_picts <- bio_picts |> 
  list_rbind()

#Saving results for later use
bio_picts |> 
  write_csv("../Outputs/average_yearly_means_picts_1985-2100.csv")

#Checking results
head(bio_picts)
```
  
**NOTE:** If you do not have access to the monthly biomass estimates in the FishMIP server, you can load the annual means we calculated in the step above and continue to run all other code chunks in this notebook.  
  
```{r}
bio_picts <- read_csv("../Outputs/average_yearly_means_picts_1985-2100.csv")
```
  
### 3. Estimating biases in annual FishMIP biomass estimates
REEFMOD provides annual biomass estimates for the GBR, so we will compare these estimates to the annual FishMIP estimates for the GBR only (step 2 above). Note that since FishMIP models only include estimates for two emission scenarios: `SSP1-2.6` and `SSP5-8.5`, we will exclude all other scenarios from these comparisons.  

#### Extracting GBR data from FishMIP models
As mentioned above, we will only use GBR biomass estimates from all FishMIP models to perform comparisons. We will then calculate ensemble minimum, mean, and maximum to calculate bias in FishMIP biomass estimates.  
  
```{r}
#Calculating ensemble statistics for GBR only
gbr_bio <- bio_picts |> 
  ungroup() |> 
  #Select GBR means
  filter(mask == 9999) |> 
  #Rename scenarios to match coral data
  mutate(scenario = case_when(scenario == "ssp126" ~ "SSP1-2.6",
                              scenario == "ssp585" ~ "SSP5-8.5",
                              T ~ scenario)) |> 
  group_by(scenario, year) |> 
  summarise(fishmip_lower_bio = min(mean_annual_bio, na.rm = T),
            fishmip_mean_bio = mean(mean_annual_bio, na.rm = T),
            fishmip_upper_bio = max(mean_annual_bio, na.rm = T))

#Checking results
head(gbr_bio)
```
    
#### Comparing FishMIP ensemble estimates to REEFMOD biomass estimates
To estimate the bias in FishMIP biomass estimates, we will divide the mean ensemble estimates for the GBR (FishMIP data) by the biomass estimates obtained from equation 2 (`fish_biomass` column in `coral` data frame).  
  
We will also obtain bias estimates and by dividing FishMIP ensemble values (minimum, mean and maximum) by the median, upper, lower biomass estimates obtained from the REEFMOD data (equations 3 to 5 stored in the `coral` data frame).  
  
```{r}
fishmip_biases <- gbr_bio |> 
  #Adding REEFMOD biomass estimates
  inner_join(coral, by = c("scenario", "year")) |> 
  #Removing coral data that is not needed
  select(!c(coral_cover, struct_complex)) |> 
  #Calculating FishMIP ensemble biases
  mutate(mean_bias_fish_bio_fishmip = fish_biomass/fishmip_mean_bio,
         lower_bias_reefmod_fishmip = reefmod_lower_biomass/fishmip_lower_bio,
         mean_bias_reefmod_fishmip = reefmod_median_biomass/fishmip_mean_bio,
         upper_bias_reefmod_fishmip = reefmod_upper_biomass/fishmip_upper_bio)

#Check results
head(fishmip_biases)
```
  
After these calculations, we have FishMIP bias estimates per year (2024-2100) and for each scenario (SSP1-2.6 and SSP5-8.5). In the next step, we will calculate a mean estimate of bias per scenario before applying this value as a correction for FishMIP biomass estimates.  
  
#### 4. Correcting FishMIP ensemble mean estimates of fish biomass
As explained in the previous step, we will not apply a bias correction that varies every year. Instead, for each scenario, we will calculate the minimum, mean, median, and maximum bias estimate. We will use these biases as corrections.  
  
```{r echo = F}
#Calculating descriptive statistics for bias estimates per scenario
mean_biases <- fishmip_biases |> 
  group_by(scenario) |> 
  #Apply calculations to biases only
  summarise(across(mean_bias_fish_bio_fishmip:upper_bias_reefmod_fishmip, 
                   #Listing statistics to be calculated
                   list(min = min, mean = mean, median = median, max = max), 
                   #Setting column names
                   .names = "{.col}-{.fn}")) |> 
  #Restructuring data frame to ease corrected biomass calculations
  pivot_longer(!scenario, names_to = "bias_type", 
               values_to = "bias_estimate") |> 
  #Separating bias type from statistic calculated
  separate_wider_delim(cols = "bias_type", delim = "-", 
                       names = c("bias_type", "stat")) |> 
  pivot_wider(names_from = bias_type, values_from = bias_estimate)

#Checking result
head(mean_biases)
```
  
Bias corrections will be applied to ensemble data, so we will calculate the minimum, mean and maximum biomass estimates per PICT across all FishMIP models.   
  
```{r}
ensemble_bio <- bio_picts |>
  #Removing GBR data
  filter(mask != 9999) |> 
   #Rename scenarios to match coral data
  mutate(scenario = case_when(scenario == "ssp126" ~ "SSP1-2.6",
                              scenario == "ssp585" ~ "SSP5-8.5",
                              T ~ scenario)) |> 
  #Calculations performed by year and EEZ
  group_by(mask, scenario, year) |> 
  #Apply calculations to biases only
  summarise(across(mean_annual_bio, 
                   #Listing statistics to be calculated
                   list(lower = min, mean = mean, max = max), 
                   #Setting column names
                   .names = "{.col}-{.fn}")) |> 
  #Restructuring data frame to ease corrected biomass calculations
  pivot_longer(!c(mask:year), names_to = "ensemble_type", 
               values_to = "ensemble_bio") |> 
  #Separating bias type from statistic calculated
  separate_wider_delim(cols = "ensemble_type", delim = "-", 
                       names = c("ensemble_type", "stat")) |> 
  pivot_wider(names_from = ensemble_type, values_from = ensemble_bio)

#Checking results
head(ensemble_bio)
```
  
REEFMOD includes data from 2024 onwards, but FishMIP data is available from 1985. This means that we do not have bias estimates before 2024, so we will use the mean bias for scenario `SSP1-2.6` to correct fish estimates for the historical period (1985-2014) and the first eight years (2015-2023) of the two emissions scenarios (SSP1-2.6 and SSP5-8.5). We chose to use the mean bias of the low emissions scenario because if mean bias from `SSP5-8.5` was used, the historical biomass values became too divergent.  
  
Finally, we produced corrected biomass from the FishMIP ensemble by applying this formula: $biomass_{est} = ensemble_{biomass} \times bias_{correction}$.  
  
```{r}
#Extracting mean biases from SSP1-2.6
mean_biases <- mean_biases |> 
  filter(stat == "mean" & scenario == "SSP1-2.6") |> 
  select(ends_with("mip")) |>
  #Matching dimension of FishMIP data to perform bias corrections
  mutate(count = nrow(ensemble_bio)) |> 
  uncount(count)

#Join FishMIP data and mean bias
bias_corr_biomass <- ensemble_bio |> 
  bind_cols(mean_biases) |> 
  #Applying corrections to biomass
  mutate(bio_corr_nash = mean_annual_bio*mean_bias_fish_bio_fishmip,
         bio_corr_reefmod_low = mean_annual_bio*lower_bias_reefmod_fishmip,
         bio_corr_reefmod_mean = mean_annual_bio*mean_bias_reefmod_fishmip,
         bio_corr_reefmod_upper = mean_annual_bio*upper_bias_reefmod_fishmip) |> 
  #Keeping only corrected columns
  select(!mean_annual_bio:upper_bias_reefmod_fishmip)

#Saving outputs
bias_corr_biomass |> 
  #Adding units for corrected biomass values
  mutate(biomass_units = "g/m2") |> 
  write_csv("../Outputs/yearly_corrected_biomass_picts_1985-2100.csv")

#Checking result
head(bias_corr_biomass)
```
  
### Plotting corrected biomass for a single PICTs
We will use biomass values corrected with the Nash derived equation from Kiribati (`mask` = 8441) for the historical period and scenario `SSP1-2.6` as an example.  
  
```{r}
bias_corr_biomass |> 
  ungroup() |> 
  filter(mask == 8441 & scenario != "SSP5-8.5") |> 
  ggplot(aes(x = year, y = bio_corr_nash, color = stat))+
  geom_line()+
  geom_point()+
  theme_bw()+
  labs(y = "Corrected fish biomass (g/m2)")+
  theme(axis.title.x = element_blank(), legend.title = element_blank())
```
  
# Calculating biomass estimates under scenario `SSP2-4.5`
FishMIP models provide biomass estimates under a low emissions (`SSP1-2.6`) and a high emissions (`SSP5-8.5`) scenarios. None of them provide biomass estimates under a moderate emissions scenario (`SSP2-4.5`). We will provide a rough estimate of biomass under `SSP2-4.5` by calculating the mean between the low (`SSP1-2.6`) and high (`SSP5-8.5`) emissions scenarios.  
  
We justify the use of this approach based on the REEFMOD data, which includes estimates of coral cover under the three scenarios of interest: `SSP1-2.6`, `SSP2-4.5`, and `SSP5-8.5`. Below, we plot coral cover under the three scenarios.  
  
```{r}
#Visualising data
p1 <- coral |> 
  #Selecting three relevant scenarios
  filter(str_detect(scenario, "2.6|4.5|8.5")) |> 
  #Plot coral cover for all scenarios
  ggplot(aes(x = year, y = coral_cover, colour = scenario, 
             linetype = scenario))+
  geom_line()+
  theme_bw()+
  labs(y = "Coral cover (%)")+
  theme(axis.title.x = element_blank(), 
        legend.position = "top", legend.direction = "horizontal")
p1
```
  
We will now calculate the mean biomass estimated under of scenarios `SSP1-2.6` and `SSP5-8.5`. We will add this to our figure as a black line to check how well it represents the moderate emissions data.  
  
```{r}
mean_26_85 <- coral |> 
  #Selecting three relevant scenarios
  filter(str_detect(scenario, "2.6|4.5|8.5")) |> 
  #Keeping only columns between year and coral cover
  select(year:coral_cover) |>
  #Reorganise data, so coral cover is now labelled by scenario
  pivot_wider(names_from = scenario, values_from = coral_cover) |> 
  #Clean up column names
  clean_names() |> 
  rowwise() |> 
  mutate(mean_scenarios = mean(c(ssp1_2_6, ssp5_8_5))) |> 
  ungroup()

p1+
  geom_line(inherit.aes = F, data = mean_26_85, aes(x = year,
                                                    y = mean_scenarios))
```
  
The averaged coral cover (black line) is closer to coral cover under scenario `SSP2-4.5` (green dashed line) than either of the two scenarios. This approach is simple, but note that before 2060, it tends to underestimate coral cover. However, after 2060 coral cover is overestimated. We expect similar biases when calculating fish biomass under the moderate emissions scenario.  
  
We will now calculate fish biomass estimates under `SSP2-4.5` for FishMIP models.   
  
```{r}
biomass_ssp245 <- bio_picts |> 
  #select data for future scenarios only
  filter(scenario != "historical") |> 
  #Reformat data before calculating mean scenario (SSP126 and SSP585)
  pivot_wider(names_from = scenario, values_from = mean_annual_bio) |> 
  rowwise() |> 
  #Calculate mean scenario
  mutate(mean_scenarios = mean(c(ssp126, ssp585))) |> 
  ungroup() |> 
  #Rename columns to match GBR data
  rename("ssp1_2_6" = "ssp126", "ssp5_8_5" = "ssp585")
  

biomass_ssp245 <- biomass_ssp245 |> 
  #Calculations performed by year and EEZ
  group_by(mask, year) |>
  summarise(across(ssp1_2_6:mean_scenarios, 
                   #Listing statistics to be calculated
                   list(min = min, median = median, max = max), 
                   #Setting column names
                   .names = "{.col}-{.fn}")) |> 
  ungroup()
```
  
We can plot the fish biomass estimates of the first four PICTs to check the results.  
  
```{r}
biomass_ssp245 |> 
  filter(mask %in% c(8318, 8316, 8317, 8315, 8324, 8314)) |> 
  #We will only plot the median values
  select(mask, year, ends_with("median")) |> 
  #Reformat data to make plotting easier
  pivot_longer(ends_with("median"), names_to = "scenario", 
               values_to = "biomass") |> 
  #Plotting
  ggplot(aes(x = year, y = biomass, colour = scenario, linetype = scenario))+
  geom_line()+
  facet_wrap(~mask, scales = "free_y")+
  theme_bw()
```
  
It should be noted that we attempted to apply corrections to the projected biomass estimates. However, results appeared to be unlikely for the moderate emissions scenario.  
  