---
title: "Biomass projections for Pacific Islands Countries and Territories (PICTs)"
author: "Beth Fulton and Denisse Fierro Arcos"
date: "2023-11-29"
output:
  github_document:
    toc: true
    html_preview: false
---

# Generating biomass projections for Pacific Islands Countries and Territories (PICTs)

The workflow described in this notebook was developed by [Dr Beth Fulton](https://orcid.org/0000-0002-5904-7917) (CSIRO) and implemented in `R` by [Denisse Fierro Arcos](https://github.com/lidefi87/). The text describing the methods was largely written by Dr Fulton with minor edits by Denisse.    
    
This notebook implements in `R` the workflow designed to generate demersal fish (and invertebrate) biomass projections for the Pacific Islands Countries and Territories (PICTs) to the year 2100. This workflow uses relationships observed in coral reefs between coral cover and fish biomass, as well as projections of coral cover in the Great Barrier Reef and fish biomass in the Pacific from the [Fisheries and Marine Ecosystem Model Intercomparison Project (Fish-MIP)](https://fish-mip.github.io/).  
  
## Loading relevant libraries
  
```{r, message = F, warning = F}
library(tidyverse)
library(openxlsx)
library(sf)
library(janitor)
library(units)
library(knitr)
```
    
## Fish-MIP models used to generate biomass projections for PICTs
To generate biomass projections for PICTs, we used biomass estimates produced by six different fisheries models:  
1. [APECOSM](https://apecosm.org/) - Apex Predators ECOSystem Model, which represents the spatialized dynamics of open ocean pelagic ecosystems in the global ocean.  
2. [BOATS](https://earthsystemdynamics.org/models/boats/) - BiOeconomic mArine Trophic Size-spectrum model simulates the global fishery as a coupled ecological-economic system.  
3. [DBPM](https://github.com/Benthic-Pelagic-Size-Spectrum-Model/dbpm_isimip_2) - The Dynamic Benthic Pelagic Model is a dynamic size spectrum model for modelling the coupling "pelagic" size-based predators and "benthic" detritivores that share a unstructured resource pool (detritus).  
4. [EcoTroph](https://doi.org/10.1016/j.ecolmodel.2009.07.031) - EcoTroph models the functioning of marine ecosystems as flows of biomass from low to high trophic levels, so as to quantify easily the impacts of fishing at an ecosystem scale.  
5. [Macroecological](https://doi.org/10.1371/journal.pone.0133794) - Macroecological is a static equilibrium model, which uses ecological and metabolic scaling theory to predict mean size composition and abundance of animals (including fish).  
6. [ZooMSS](https://doi.org/10.1016/j.ecolmodel.2020.109265) - The Zooplankton Model of Size Spectra is a functional size-spectrum model of the marine ecosystem to resolve phytoplankton, nine zooplankton functional groups (heterotrophic flagellates and ciliates, omnivorous and carnivorous copepods, larvaceans, euphausiids, salps, chaetognaths and jellyfish) and three size-based fish groups.  
    
## 1. Biomass projections from REEFMOD data
A relationship between coral cover and fish biomass has been established in the literature. A quick synthesis of existing relationships can be found in the document `Relationships for fish biomass model.docx`.  

### Relationships between coral cover and fish biomass
The most useful relationships came from the [Graham and Nash (2013)](https://doi.org/10.1007/s00338-012-0984-y) data set (explored in `graham_nash_2012_dataset.xlsx`). Using the raw data gives the following relationships:

**Equation 1**: Structural complexity as a function of coral cover. ($r^2$ = 0.4314)  
$struct_{complexity} = 3e^{-7} \times {coral_{cover}}^4 - 5e^{-5} \times {coral_{cover}}^3 + 0.0022 \times {coral_{cover}}^2 + 1.3892$  
  
```{r}
#Base spreadsheet
working_data <- "../Analysis/Excel Sheets/bias_correction_calculations_mean_coral_cover_all_climate_scenarios.xlsx"

#Loading coral cover data
coral <- read.xlsx(working_data, sheet = 1)

#Calculating structural complexity using equation 1 above
coral <- coral |> 
  #rearrange data to facilitate calculations
  pivot_longer(-Year, names_to = "scenario", values_to = "coral_cover") |> 
  #make names small case
  clean_names() |> 
  #calculate structural complexity
  mutate(struct_complex = (3e-7*(coral_cover^4))-((5e-5)*(coral_cover^3))+(0.0022*(coral_cover^2))+1.3892)

#Checking results
head(coral)
```
  
```{r}
#Visualising data
coral |> 
  ggplot(aes(x = year, y = coral_cover, colour = scenario, linetype = scenario))+
  geom_line()+
  theme_bw()+
  labs(y = "Coral cover (%)")+
  theme(axis.title.x = element_blank(), 
        legend.position = "top", legend.direction = "horizontal")
```
  
**Equation 2**: Fish biomass ($kg \times ha^{-1}$) as a function of structural complexity ($r^2$ = 0.6102). Note that a logistic is a little poorer fit, so using a quadratic on the argument that once a reef habitat is too complex, it loses places for fish to sit.  
$fish_{biomass} = -2294.6 \times {struct_{complexity}}^2 + 8961.1 \times struct_{complexity} - 6843.6$  
  
In this step, we will also convert fish biomass from $kg \times ha^{-1}$ to $g \times m^{-2}$, so it matches the outputs of Fish-MIP models.  
  
```{r}
#Calculating fish biomass
coral <- coral |> 
  mutate(fish_biomass_kg_ha = (-2294.6*(struct_complex^2))+(8961.1*struct_complex)-6843.6) |> 
  mutate(fish_biomass = fish_biomass_kg_ha*1000/10000)

#Checking results
head(coral)
```
  
The REEF_MOD project (of Yves-Marie Bozec and Peter Mumby at University of Queensland) has postulated a slightly different set of relationships from the same data (basically a linear relationship).  
  
Fitting a line to the lower, median and upper bounds of this relationship gives the following equations:  
  
**Equation 3**: REEFMOD lower bound equation: $fish_{biomass} = 12.716 \times coral_{cover} + 146.75$  
**Equation 4**: REEFMOD median equation: $fish_{biomass} = 13.56 \times coral_{cover} + 732.15$  
**Equation 5**: REEFMOD upper bound equation: $fish_{biomass} = 14.285 \times coral_{cover} + 1325.7$  
  
```{r}
#Calculating REEFMOD biomass values
coral <- coral |> 
  #lower biomass bound
  mutate(reefmod_lower_biomass = (12.716*coral_cover+146.75)*1000/10000,
         #median biomass
         reefmod_median_biomass = (13.56*coral_cover+732.15)*1000/10000,
         #upper biomass bound
         reefmod_upper_biomass = (14.285*coral_cover+1325.7)*1000/10000) |> 
  #We will drop the biomass in Kg/ha
  select(!fish_biomass_kg_ha)

head(coral)
```
  
Note that the REEFMOD team predicted mean coral coverage from 2024 to 2100 for the Great Barrier Reef (GBR) under five emissions scenarios: `SSP1-1.9`, `SSP1-2.6`, `SSP2-4.5`, `SSP3-7.0`, and `SSP5-8.5`. The above calculations were applied to these five projections.    
  
## 2. Biomass projections from Fish-MIP models
Monthly projected biomass for the GBR under scenarios `SSP1-2.6` and `SSP5-8.5` were extracted from Fish-MIP global models using a polygon. Annual time series were calculated for Fish-MIP models forced by both GFDL and IPSL general circulation models.  
  
Projections from DBEM were removed from further analysis because biomass values were 100x lower than all other models. Similarly, biomass values projected by Ecotroph forced by GFDL were removed as those values were an order of magnitude higher than all other models.  
  
Biomass projections from all other models were used to calculate a biomass ensemble minimum, mean and maximum from 2024 to 2100.  
  
### Extracting biomass data for the GBR
We will extract biomass estimates for the GBR from all global Fish-MIP models using a mask derived from the GBR outer boundaries. We will then calculate mean yearly biomass for each model and scenario.  
  
```{r}
#Folder containing outputs from Fish-MIP models
base_folder <- "/rd/gem/private/users/camillan/Extract_tcblog10_Data/Output/sumSize_annual/sizeConsidered10g_10kg/EEZsummaries/gridded_outputs/"
#Listing all relevant files to calculate biomass projections
global_files <- list.files(base_folder, full.names = T)

#Loading gbr mask
gbr_mask <- read_csv("../Data/GBR_Outer_Boundary/GBR_outer_boundary_mask.csv")

#Defining function to extract GBR biomass data from Fish-MIP outputs
mean_yr_bio <- function(file_name, mask){
  #Load file
  da <- read.csv(file_name) |> 
    #Extract GBR data - by applying mask
    right_join(mask, by = c("x", "y")) |> 
    #Removing columns not needed
    select(!area_m:layer) |> 
    #Calculating mean yearly biomass
    group_by(year, mem, esm, scenario) |> 
    summarise(mean_annual_bio = mean(biomass, na.rm = T))
}

#Apply function to all Fish-MIP output files
bio_data <- global_files |> 
  map(\(x) mean_yr_bio(x, gbr_mask))

#Combine all list elements into a single data frame
bio_data <- bio_data |> 
  list_rbind()

#Checking results
head(bio_data)
```
    
### Calculating biomass ensemble statistics
We will calculate minimum, mean and maximum biomass for the entire ensemble between 2024 and 2100.  
  
```{r}
#Calculating ensemble statistics
bio_data <- bio_data |> 
  ungroup() |> 
  #Rename scenarios to match coral data
  mutate(scenario = case_when(scenario == "ssp126" ~ "SSP1-2.6",
                              scenario == "ssp585" ~ "SSP5-8.5",
                              T ~ scenario)) |> 
  group_by(scenario, year) |> 
  summarise(fishmip_lower_bio = min(mean_annual_bio, na.rm = T),
            fishmip_mean_bio = mean(mean_annual_bio, na.rm = T),
            fishmip_upper_bio = max(mean_annual_bio, na.rm = T))

#Checking results
head(bio_data)
```
    
## 3. Comparing Fish-MIP and REEFMOD biomass estimates
For scenarios `SSP1-2.6` and `SSP5-8.5`, we will compare the biomass estimates calculated from REEFMOD data (step 1) and the Fish-MIP model ensemble (step 2). We will divide the ensemble mean by the biomass estimates obtained from equation 2 (`fish_biomass` column in `coral` data frame), and by the median, upper, lower bounds obtained from REEFMOD data (equations 3 to 5). This calculations will help us estimate the bias of the Fish-MIP global model ensemble.  
  
```{r}
fishmip_biases <- bio_data |> 
  #Adding REEFMOD biomass estimates
  inner_join(coral, by = c("scenario", "year")) |> 
  #Removing coral data that is not needed
  select(!c(coral_cover, struct_complex)) |> 
  #Calculating Fish-MIP ensemble biases
  mutate(mean_bias_fish_bio_fishmip = fish_biomass/fishmip_mean_bio,
         lower_bias_reefmod_fishmip = reefmod_lower_biomass/fishmip_lower_bio,
         mean_bias_reefmod_fishmip = reefmod_median_biomass/fishmip_mean_bio,
         upper_bias_reefmod_fishmip = reefmod_upper_biomass/fishmip_upper_bio)
  
#Check results
head(fishmip_biases)
```
  
## 4. Calculating proportion of demersal biomass (*optional step*)
Two Fish-MIP models provide demersal and pelagic biomass estimates: FEISTY (forced by GFDL model) and DBPM (forced by IPSL model). We will use this information to calculate the ratio of the demersal component from the total consumer biomass (demersal + pelagic biomass).  
  
```{r}
#Load monthly biomass
bio_data_all_models <- read.xlsx(working_data, sheet = "vsFISHMIP All TimeSteps", startRow = 2, detectDates = T)

#We will divide this scenario
scenarios <- which(str_detect(names(bio_data_all_models), "SSP"))

#Creating empty data frame to save results
dem_pel_bio <- data.frame()
#Create new longer data frame - Split at each scenario
for(i in 1:length(scenarios)){
  #Find the column with scenario name
  start <- scenarios[i]
  #End in the column before next scenario
  end <- scenarios[i+1]-1
  #If it is the last scenario, select everything until the end
  if(is.na(end)){
    end <- ncol(bio_data_all_models)
  }
  #Get scenario name
  ssp <- names(bio_data_all_models)[start]
  #Extract data between start and end columns
  da <- bio_data_all_models[,(start+1):end] |> 
    #Reshaping data
    pivot_longer(!Year, names_to = "fish_model", values_to = "proj_biomass") |> 
    #Add scenario
    mutate(scenario = ssp,
           #Get year from date
           year = year(Year),
           #Get month from date
           month = month(Year)) |> 
    #Remove date (incorrectly labelled as Year)
    select(!Year) |> 
    #Remove data before 2024
    filter(year >= 2024) |> 
    #Extracting demersal and pelagic biomass from FEISTY and DBPM models to be later
    filter(str_detect(fish_model, "_DEM|_PELAG")) |> 
    #Remove FEISTY IPSL because it is not needed in this calculation
    filter(str_detect(fish_model, "FEISTY_IPSL", negate = T))
  #Putting everything together in new data frame
  dem_pel_bio  <- dem_pel_bio  |> 
    bind_rows(da)
}

#Calculate proportion of demersal biomass
dem_prop <- dem_pel_bio |> 
  #Remove models with no data
  drop_na(proj_biomass) |>
  #Calculate mean biomass by year, model and scenario
  group_by(scenario, fish_model, year) |> 
  summarise(mean_yr_biomass = mean(proj_biomass, na.rm = T)) |> 
  #Adding column with biomass type
  mutate(bio_type = str_split_i(fish_model, "_", i = -1),
         fish_model = str_remove(fish_model, "_DEM|_PELAG")) |> 
  #Rearranging table to calculate demersal biomass proportions
  pivot_wider(names_from = bio_type, values_from = mean_yr_biomass) |> 
  #Calculating proportion
  mutate(prop_dem = DEM/(DEM+PELAG))

#Checking result
head(dem_prop)
```
  
## 5. Bias correcting Fish-MIP ensemble biomass projections
The bias estimates and demersal proportions calculated above will be used into calculate the biomass per hectare for each PICT for two scenarios: `SSP1-2.6` and `SSP5-8.5`. Biomass estimates from Fish-MIP global models were extracted using the EEZ boundaries of each PICT. This information was used to calculate the minimum, mean and maximum biomass estimates for the entire ensemble. Note that the same models excluded in step 3 are also excluded here. Finally, a set of bias corrected time series are then generated based on the following combinations:
  
```{r echo = F}
kable(data.frame(vs = c("Emsemble min", "Emsemble mean", "Emsemble max"),
                 bias_based_on_Nash = rep("Biomass estimates (eq 2)", 3),
                 bias_based_on_min = rep("REEFMOD min (eq 3)", 3),
                 bias_based_on_mean = rep("REEFMOD median (eq 4)", 3),
                 bias_based_on_max = rep("REEFMOD max (eq 5)", 3)))
```
  
In each case the biomass is given as:  
  
$biomass_{est} = ensemble_{biomass} \times bias_{correction}$  
  






