---
title: "Plotting biomass projections for PICTs"
author: "Denisse Fierro Arcos"
date: "2023-12-13"
output:
  github_document:
    toc: true
    html_preview: false
---

## Loading libraries
    
```{r, message = F, warning = F}
library(readr)
library(dplyr)
library(ggplot2)
library(stringr)
library(tidyr)
```
  
## Loading biomass data from FishMIP models
  
```{r}
#Loading keys identifying EEZs
PICTS_keys <- read_csv("../Outputs/SouthPacific_EEZ-GBR_keys.csv") |> 
  #Removing original ID - using simplified version
  select(!MRGID) |> 
  #Remove duplicates
  distinct() |> 
  #Add region information
  left_join(read_csv("../Data/PICT_regions.csv"), 
            by = join_by("name"))
  
#Loading FishMIP biomass data
bio_picts <- read_csv("../Outputs/average_yearly_means_picts_1985-2100.csv")|>
  #Removing GBR data
  filter(mask != 9999)

#Calculating SSP2-4.5 biomass estimates
bio_picts_ssp245 <- bio_picts |> 
  #select data for future scenarios only
  filter(scenario != "historical") |> 
  #Reformat data before calculating mean scenario (SSP126 and SSP585)
  pivot_wider(names_from = scenario, values_from = mean_annual_bio) |> 
  rowwise() |> 
  #Calculate mean scenario
  mutate(ssp245_est = mean(c(ssp126, ssp585))) |> 
  ungroup() |> 
  #Removing columns not needed
  select(!(ssp126:ssp585)) |> 
  #Rearranging data before merging to original FishMIP scenarios
  pivot_longer(ssp245_est, names_to = "scenario", values_to = "mean_annual_bio") |> 
  #Ensuring column locations match
  relocate(mask, .after = scenario) |> 
  #Merging data frames
  bind_rows(bio_picts) |> 
  #Rename scenarios to match coral data
  mutate(scenario = case_when(scenario == "ssp126" ~ "SSP1-2.6",
                              scenario == "ssp245_est" ~ "SSP2-4.5",
                              scenario == "ssp585" ~ "SSP5-8.5",
                              T ~ scenario))
```
  
## Calculating mean for reference period
Plots created here will show percentage change in biomass over time under two emissions scenarios rather than absolute values. Our reference period is between 2010 and 2020. From 2010 to 2014, biomass from the historical period will be used, and between 2015 and 2020 we will use `SSP1-2.6` data as this is the best case scenario.  

```{r}
#Calculating mean for reference decade (2010-2020)
ref_decade <- bio_picts |>
  filter(year >= 2010 & year <= 2020) |> 
  filter(str_detect(scenario, "585", negate = T)) |> 
  group_by(mem, esm, mask) |> 
  summarise(ref = mean(mean_annual_bio))

#Calculating proportion of biomass in relation to reference decade
bio_plots <- bio_picts_ssp245 |> 
  left_join(ref_decade, by = c("mem", "esm", "mask")) |> 
  mutate(rel_change = ((mean_annual_bio-ref)/ref)*100) |> 
  #Calculations performed by year and EEZ
  group_by(mask, scenario, year) |> 
  #Apply calculations to biases only
  summarise(across(rel_change, 
                   #Listing statistics to be calculated
                   list(lower = min, median = median, max = max), 
                   #Setting column names
                   .names = "{.col}_{.fn}")) |> 
  ungroup() |> 
  left_join(PICTS_keys, by = c("mask"="ID"))

#Saving results
bio_plots |> 
  write_csv("../Outputs/yearly_relative_biomass_change_1985-2100.csv")

#Checking results
head(bio_plots)
```
  
## Plotting data 
We will remove the lowest emission scenario (`SSP1-2.6`) before plotting relative change in biomass. We will create three plots based on the three cultural regions found in this area: Melanesia, Micronesia, and Polynesia. All plots will follow the same format, so we will define a base plot and apply it to all regions.    
  
```{r}
base_gg <- list(geom_line(linewidth = 0.5),
                #Horizontal line at 0 as reference for no change
                geom_hline(yintercept = 0, color = "#709fcc", 
                           linewidth = 0.65, linetype = 2),
                #Vertical line showing reference between historical and future
                #scenarios
                geom_vline(xintercept = 2015, color = "#709fcc", 
                           linewidth = 0.65),
                #Add ribbon (min and max)
                geom_ribbon(aes(ymin = rel_change_lower, ymax = rel_change_max, 
                                fill = scenario), alpha = 0.3, color = NA),
                #Change x scale to make it easier to identify years
                scale_x_continuous(minor_breaks = seq(1980, 2100, by = 10),
                                   breaks = seq(1980, 2100, by = 40), 
                                   limits = c(1980, 2100)),
                #Make individual plots by PICT
                facet_wrap(~name, scales = "free_y"),
                #Change line colours 
                scale_color_manual(values = c("historical" = "black", 
                                              "SSP2-4.5" = "#33bbee", 
                                              "SSP5-8.5" = "#ee3377")),
                #Change ribbon colours
                scale_fill_manual(values = c("historical" = "black", 
                                             "SSP2-4.5" = "#33bbee", 
                                             "SSP5-8.5" = "#ee3377")),
                #Ensure legend has three rows
                guides(color = guide_legend(ncol = 3, label.hjust = 0.5)),
                #Apply pre-defined theme
                theme_bw(),
                #Final changes to make plot more appealing and easy to read
                theme(legend.position = "top", 
                      legend.key.size = unit(1., "cm"),
                      panel.grid.minor.y = element_blank(),
                      plot.margin = margin(b = 1.15, r = 0.5, l = 0.5, t = 0.2, 
                                           unit = "cm"),
                      legend.text = element_text(size = 12), 
                      strip.text = element_text(size = 12), 
                      axis.title.x = element_blank(),
                      legend.title = element_blank(), 
                      axis.title.y = element_text(size = 12),
                      axis.text.x = element_text(angle = 45, vjust = 0.765, 
                                                 hjust = 0.65)),
                #Chaning labels in y axis
                ylab("% biomass change (reference period: 2010-2020)"))
```
  
### Melanesia plot
  
```{r}
bio_plots |> 
  #Removing lowest emission scenario
  filter(str_detect(scenario, "2.6", negate = T) & Region == "Melanesia") |> 
  ggplot(aes(x = year, y = rel_change_median, color = scenario))+
  base_gg
```
    
```{r, eval = F}
ggsave("../Outputs/melanesia_relative_biomass_change.pdf", device = "pdf", 
       width = 12, height = 6)
```
  
### Micronesia plot
  
```{r}
bio_plots |> 
  #Removing lowest emission scenario
  filter(str_detect(scenario, "2.6", negate = T) & Region == "Micronesia") |> 
  ggplot(aes(x = year, y = rel_change_median, color = scenario))+
  base_gg
```
    
```{r, eval = F}
ggsave("../Outputs/micronesia_relative_biomass_change.pdf", device = "pdf", 
       width = 12, height = 8)
```
  
### Polynesia plot
  
```{r}
bio_plots |> 
  #Removing lowest emission scenario
  filter(str_detect(scenario, "2.6", negate = T) & Region == "Polynesia") |> 
  ggplot(aes(x = year, y = rel_change_median, color = scenario))+
  base_gg
```
    
```{r, eval = F}
ggsave("../Outputs/polynesia_relative_biomass_change.pdf", device = "pdf", 
       width = 12, height = 8)
```
  