---
title: "Mapping biomass change under two emissions scenarios for PICTs"
author: "Denisse Fierro Arcos"
date: "2024-01-03"
output:
  github_document:
    toc: true
    html_preview: false
---

# Loading libraries
  
```{r, message = F, warning = F}
library(data.table)
library(tidyverse)
library(rnaturalearth)
library(cmocean)
library(cowplot)
library(sf)
library(nngeo)
```

# Loading information about PICTs
This dataset will allow us to identify the country associated to each unique ID in the fish biomass estimates from FishMIP models. We will exclude the Great Barrier Reef (GBR) because it is not relevant here.  
  
```{r}
PICTS_keys <- read_csv("../Outputs/SouthPacific_EEZ-GBR_keys.csv") |> 
  #Removing GBR data
  filter(name != "GBR")
```
  
# Loading fish biomass estimates from FishMIP models
We will create maps that show the percentage change in fish biomass under two scenarios: `SSP2-4.5` and `SSP5-8.5` for the decades between 2045 and 2055, and between 2085 and 2095. First, we will identify all files containing relevant information and then we will extract the data we need to create our maps.    
  
```{r}
#Folder containing outputs from FishMIP models
base_folder <- "/rd/gem/private/users/camillan/Extract_tcblog10_Data/Output/sumSize_annual/sizeConsidered10g_10kg/EEZsummaries/gridded_outputs/"
#Listing all relevant files to calculate biomass projections
global_files <- list.files(base_folder, full.names = T)
#Models
members <- str_extract(global_files, "outputs//(.*)_(h|s)", group = 1) |> 
  unique()
```
  
We will go through each file, extract biomass estimates for `SSP1-2.6` and `SSP5-8.5` to estimate biomass under a moderate emissions scenario (`SSP2-4.5`) because FishMIP models do not include this scenario. As explained in the [04_Biomass_projection_SouthPacific](https://github.com/Fish-MIP/Extract_PICTs/blob/main/Scripts/04_Biomass_projections_SouthPacific.md) notebook, `SSP2-4.5` biomass estimates are calculated by taking the mean of the lowest and highest emissions scenarios (`SSP1-2.6` and `SSP5-8.5`).  
  
We should note that this approach provides a rough estimate of fish biomass under `SSP2-4.5`, which may result in an under estimate prior to 2060, and an overestimate after 2080.  
  
```{r, evaluate = F}
for(m in members){
  #Load all data available for a single FishMIP model
  df_model <- str_subset(global_files, m) |> 
    #Ignore columns SOVEREIGN1-3 - not needed here
    map_df(~fread(., drop = c(paste0("SOVEREIGN", 1:3), "area_m"))) |> 
    #Extract data only for years to be used in maps
    filter(year >= 2010 & year <= 2020 | year >= 2045 & year <= 2055 | year >= 2085 & year <= 2095) |> 
    #Do not keep data before 2021 for scenario ssp585
    filter(!((year >= 2015 & year <= 2020) & scenario == "ssp585")) |> 
    #If EEZ is not a PICT mark as NA
    mutate(eez = case_when(!eez %in% PICTS_keys$MRGID ~ NA,
                           T ~ eez)) |> 
    #Remove EEZs classified as NA (no PICTs)
    drop_na(eez) 
  
  #Calculating biomass for SSP2-4.5 scenario
  ssp245 <- df_model |> 
    #Keeping data from 2045 and beyond for two scenarios
    filter(year >= 2045) |> 
    #Calculating mean at a pixel level
    group_by(x, y, year, mem, esm) |>
    summarise(biomass = mean(biomass, na.rm = F)) |> 
    #Reformatting data so it matches original file
    relocate(biomass, .after = "year") |> 
    #Adding scenario name
    mutate(scenario = "ssp245") |> 
    #Adding missing EEZ and name information
    left_join(df_model |> 
                distinct(x, y, eez, GEONAME), by = c("x", "y"))
  
  #We will now remove the SSP1-2.6 data from 2045 and beyond 
  df_model <- df_model |> 
    filter(!(scenario == "ssp126" & year >= 2045)) |> 
    #This will be replaced with SSP2-4.5
    bind_rows(ssp245) |> 
    #Create new group column to calculate means
    mutate(group = case_when(year <= 2020 ~ "reference",
                             year >= 2045 & year <= 2055 ~ "mean50",
                             year >= 2085 & year <= 2095 ~ "mean80"),
           #The mean50 and mean80 groups also need to have the scenario as part of the label
           group = case_when(group != "reference" ~ str_c(group, scenario, sep = "_"),
                             T ~ group)) |> 
    #Calculate mean per ensemble member
    group_by(x, y, mem, esm, eez, GEONAME, group) |> 
    summarise(mean_bio = mean(biomass, na.rm = T)) |> 
    #Reorganise table to facilitate calculations
    pivot_wider(names_from = group, values_from = mean_bio) |> 
    ungroup() |> 
    #Calculate % change in fish biomass for two scenarios: SSP2-4.5 and SSP5-8.5
    mutate(rel_change_mean50_ssp245 = ((mean50_ssp245-reference)/reference)*100,
           rel_change_mean50_ssp585 = ((mean50_ssp585-reference)/reference)*100,
           rel_change_mean80_ssp245 = ((mean80_ssp245-reference)/reference)*100,
           rel_change_mean80_ssp585 = ((mean80_ssp585-reference)/reference)*100)
    
  #Create name to save file  
  f_out <- file.path("../Outputs", str_c(m, "_map_data.csv"))

  #Saving results for each model
  df_model |> 
    fwrite(f_out)
}
```
    
# Calculating % change in fish biomass for FishMIP ensemble
In the previous step, decadal % change in fish biomass was calculated for each FishMIP ensemble member. Now, we will calculate the mean change for the entire ensemble.
  
```{r}
#Load grid sample
mask_base <- read_csv("../Outputs/mask_1deg.csv") |> 
  select(!mask) |> 
  rename(x = Lon, y = Lat)

#Listing all relevant files to calculate biomass projections
maps_data <- list.files("../Outputs/", pattern = "_map_data.csv", full.names = T) |> 
  map_df(~fread(.)) |> 
  #Calculations performed by year and EEZ
  group_by(x, y, eez, GEONAME) |> 
  #Apply calculations to biases only
  summarise(across(reference:rel_change_mean80_ssp585, 
                   #Listing statistics to be calculated
                   list(median = median), 
                   #Setting column names
                   .names = "{.col}_{.fn}")) |> 
  ungroup() |> 
  #Change longitudes range to 0-360 degrees
  mutate(x = x%%360) 
```
  
# Loading relevant shapefiles
We will use a map of the world as a base map, and the EEZ boundaries of PICTs.  
  
## Base map
World map to be used as base map.  
  
```{r}
#Base map
world <- ne_countries(returnclass = "sf") |> 
  st_shift_longitude() |>
  filter(continent %in% c("Oceania", "Asia"))
```
  
## PICT EEZ boundaries
Since the original boundaries have longitude between $-180^{\circ}$ and $180^{\circ}$, we will change it to $0-360^{\circ}$ so all EEZs show in the centre of the map.  
  
```{r}
picts <- read_sf("../Outputs/SouthPacific_EEZ-GBR.shp") |> 
  #Exclude Great Barrier Reef
  filter(name != "GBR") |> 
  #Fix any invalid geometry
  st_make_valid() |> 
  #Dissolve boundaries within PICTs
  group_by(ID_1, name) |> 
  summarise(across(geometry, ~ st_combine(.))) |> 
  #Ensure ID and name are factors
  mutate(ID_1 = as.factor(ID_1)) |> 
  #Ensuring shapefile crosses the international dateline
  st_shift_longitude() |> 
  #Removing holes from islands inside EEZ
  st_remove_holes()
```
  
# Plotting maps
In this section, we will plot individual maps for each decade/scenario combinations. Then, we will merge them together in a single plot.  
  
## Creating colour palette
We will create colour palette that is suited to the data we are plotting. Given that the range of negative change is larger than the positive change, this colour palette will have a much larger range of colours for negative values.  
  
```{r}
pal <- c(cmocean("matter", start = 0.1, end = 0.8, direction = -1)(179), 
         cmocean("delta", start = 0.49, end = 0.5)(20),
         cmocean("deep", start = 0.1, end = 0.3)(57))
```
  
## Creating individual maps
All maps share the same colour palette, so we will not include legends in any of them except for the last one.  
  
Grey cells are areas with no data in the models.  
  
```{r}
#SSP2-4.5 2045-2055 
p50_245 <- maps_data |>
  ggplot(aes(x, y, fill = rel_change_mean50_ssp245_median))+
  geom_tile()+
  scale_fill_stepsn(colors = pal, n.breaks = 11, 
                    limits = c(-75, 15), show.limits = T)+
  geom_sf(inherit.aes = F, data = picts, fill = NA, 
          show.legend = F, linewidth = 0.25)+
  geom_sf(inherit.aes = F, data = world)+
  theme_bw()+
  labs(title = "SSP2-4.5: 2045-2055")+
  theme(axis.title = element_blank(), 
        plot.title = element_text(hjust = 0.5),
        legend.position = "none")+
  lims(x = c(110, 240), y = c(-45, 25))

#SSP5-8.5 2045-2055 
p50_585 <- maps_data |> 
  ggplot(aes(x, y, fill = rel_change_mean50_ssp585_median))+
  geom_tile()+
  scale_fill_stepsn(colors = pal, n.breaks = 11, 
                    limits = c(-75, 15), show.limits = T)+
  geom_sf(inherit.aes = F, data = picts, fill = NA, 
          show.legend = F, linewidth = 0.25)+
  geom_sf(inherit.aes = F, data = world)+
  theme_bw()+
  labs(title = "SSP5-8.5: 2045-2055")+
  theme(axis.title = element_blank(), 
        plot.title = element_text(hjust = 0.5),
        legend.position = "none")+
  lims(x = c(110, 240), y = c(-45, 25))

#SSP2-4.5 2085-2095 
p80_245 <- maps_data |> 
  ggplot(aes(x, y, fill = rel_change_mean80_ssp245_median))+
  geom_tile()+
  scale_fill_stepsn(colors = pal, n.breaks = 11, 
                    limits = c(-75, 15), show.limits = T)+
  geom_sf(inherit.aes = F, data = picts, fill = NA,
          show.legend = F, linewidth = 0.25)+
  geom_sf(inherit.aes = F, data = world)+
  theme_bw()+
  labs(title = "SSP2-4.5: 2085-2095")+
  theme(axis.title = element_blank(), 
        plot.title = element_text(hjust = 0.5),
        legend.position = "none")+
  lims(x = c(110, 240), y = c(-45, 25))

#SSP5-8.5 2085-2095 
p80_585 <- maps_data |> 
  ggplot(aes(x, y, fill = rel_change_mean80_ssp585_median))+
  geom_tile()+
  scale_fill_stepsn(colors = pal, n.breaks = 11, 
                    limits = c(-75, 15), show.limits = T)+
  geom_sf(inherit.aes = F, data = picts, fill = NA, 
          show.legend = F, linewidth = 0.25)+
  geom_sf(inherit.aes = F, data = world)+
  theme_bw()+
  guides(fill = guide_legend(title = "% change", title.position = "top", 
                             title.hjust = 0.5, 
                             label.position = "bottom", nrow = 1))+
  labs(title = "SSP5-8.5: 2085-2095")+
  theme(axis.title = element_blank(), plot.title = element_text(hjust = 0.5), 
        legend.spacing.x = unit(0, "cm"), legend.position = "bottom", 
        legend.direction = "horizontal", legend.key.width = unit(1.5, "cm"), 
        legend.key.height = unit(0.75, "cm"))+
  lims(x = c(110, 240), y = c(-45, 25))

#Get legend
legend <- get_legend(p80_585)

#Remove legend from last plot
p80_585 <- p80_585+theme(legend.position = "none")

#Create title for 
title <- ggdraw()+
  draw_label("Mean % change in fish biomass from 2010-2020")+
  theme(plot.margin = margin(0, 0, 0, 0, unit = "cm"))

#Plotting everything together
all_plots <- plot_grid(title, plot_grid(p50_245, p50_585, ncol = 2, nrow = 1),
                       plot_grid(p80_245, p80_585, ncol = 2, nrow = 1),
                       legend, ncol = 1, nrow = 4, 
                       rel_heights = c(0.15, 1, 1, 0.4))

#Check final map
all_plots
```
  
## Saving final plot
  
```{r, eval = F}
ggsave("../Outputs/maps_perc_change_PICTs_40s-80s.pdf", device = "pdf", width = 14, height = 9)
```

